### 2013-07-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
document formatting for appearance

%%---------------------------------------------------------------------------%%
\section{Derivation of the Monoenergetic $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations as derived in Appendix~\ref{chap:pn_equations}
give $N+1$ coupled first-order equations capturing the spatial and
angular-dependence of the solution. In multiple dimensions, the
equation set becomes large and coupled not only through angular
moments but also through the spatial variables. As a simpler
alternative to multidimensional $P_N$ solutions, Gelbard recognized in
1960 that the planar $P_N$ equations could be simplified and applied
an ad-hoc method to extend them to multiple dimensions, yielding the
$SP_N$ equations. These equations are not only fewer in number, but
also take on a diffusion-like form while maintaining the angular
character of the flux, making them amenable to solutions with modern
diffusion methods.

@@

yields the $SP_3$ equations). Boundary conditions for these equations
are provided in Appendix~\ref{chap:spn_boundary_conditions}, the
multigroup form is presented in Appendix~\ref{chap:mg_spn_equations},
and the spatial discretization in
Appendix~\ref{chap:spn_spatial_discretization}.

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
source derived in \S~\ref{sec:pn_eigenvalue}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}
We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
at preconditioning this system will be to use the point Jacobi
preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
to this diagonal form.

@@

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of
MCSA. To investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. We
therefore require a more expansive set of preconditioning techniques
to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
which MCSA is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

Figure~\ref{fig:ilut_iterations} gives the number of iterations to
converge the fixed source problem to a tolerance of \sn{1}{8}. As
expected, the higher the fill level chosen for the ILUT factorization,
the fewer iterations are required to converge the problem (only 9 MCSA
iterations were required to converge the problem for the fill level of
3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
the iterations needed to converge were not as sensitive, signaling
that a larger drop tolerance can perhaps be used without a significant
degradation in iterative performance. At smaller fill levels, the
sensitivity to the ILUT drop tolerance is more significant. For all
fill levels used, convergence was not achieved for the fuel assembly
problem with a drop tolerance larger than \sn{1}{-3}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge an
      eigenvalue iteration for the fuel assembly problem with ILUT
      preconditioning as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the iteration behavior for a
      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
      were used.}}
  \label{fig:ilut_iterations}
\end{figure}

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. Second, the matrix-matrix multiply operations required to
build the composite operator see significant performance losses due to
the amount of data that must be handled. In parallel, losing sparsity
severely inhibits performance where now parallel operations require
communications among orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.} \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
among states in the system and reduce the amount of memory used
along with the potential latency overhead. Given these two objectives,
the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge a
      single eigenvalue iteration for the fuel assembly problem with
      each preconditioning as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/rda_quality.pdf}
  \end{center}
  \caption{\textbf{Preconditioning quality metric for the fuel
      assembly problem given as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_quality}
\end{figure}

@@

In this chapter, the Monte Carlo synthetic acceleration method has
been applied to the $SP_N$ discretization of the neutron transport
equation and explored in the context of a difficult nuclear fuel
assembly criticality problem. The following are the significant
observations and findings.

>>>>

%%---------------------------------------------------------------------------%%
\section{Derivation of the Monoenergetic $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations as derived in \citep{lewis_computational_1993}
give $N+1$ coupled first-order equations capturing the spatial and
angular-dependence of the solution:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
dimensions, the equation set becomes large and coupled not only
through angular moments but also through the spatial variables. As a
simpler alternative to multidimensional $P_N$ solutions, Gelbard
recognized in 1960 that the planar $P_N$ equations could be simplified
and applied an ad-hoc method to extend them to multiple dimensions,
yielding the $SP_N$ equations. These equations are not only fewer in
number, but also take on a diffusion-like form while maintaining the
angular character of the flux, making them amenable to solutions with
modern diffusion methods.

@@

yields the $SP_3$ equations).

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
\citep{evans_simpli_2013}. We again apply the change of variables to
yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}
We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
at preconditioning this system will be to use the point Jacobi
preconditioning from \S\ref{subsubsec:basic_mcsa_preconditioning} due
to this diagonal form.

@@

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of
MCSA. To investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing as a function of spectral radius. Near a spectral radius of
0.97, the number of histories required to converge MCSA in less than
100 iterations takes a dramatic rise that exhibits neither exponential
nor power law behavior. As the spectral radius approaches 1, the
number of histories required becomes significant and effectively
impractical to compute. Even with this simple diffusion problem, the
behavior is consistent with that observed for the fuel assembly
problem with $SP_1$ discretization. In that case, we estimated a
spectral radius of $\approx 0.998$, larger than any of the spectral
radii that could be computed within even 90 minutes of compute time
for this simple two dimensional problem. For that problem, even
1,000,000 histories ($\approx 50$ per DOF) was not enough to provide
convergence. Even if single solve times of an hour can be tolerated,
dozens of solves are typically required to compute the multiplication
factor and flux spectrum within the $SP_N$ eigenvalue scheme for more
difficult problems like the fuel assembly, making the method unusable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories per DOF required to converge as a
      function of spectral radius for the neutron diffusion problem.}
    \textit{The number of histories was increased to achieve
      convergence in less than 100 iterations. At least 10,000
      histories were used for each calculation with 10,000 DOFs in the
      problem.}}
  \label{fig:breakdown_histories}
\end{figure}

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. We
therefore require a more expansive set of preconditioning techniques
to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
which MCSA is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

Figure~\ref{fig:ilut_iterations} gives the number of iterations to
converge the fixed source problem to a tolerance of \sn{1}{8}. As
expected, the higher the fill level chosen for the ILUT factorization,
the fewer iterations are required to converge the problem (only 9 MCSA
iterations were required to converge the problem for the fill level of
3 and drop tolerance of \sn{1}{-5} case). For higher levels of fill,
the iterations needed to converge were not as sensitive, signaling
that a larger drop tolerance can perhaps be used without a significant
degradation in iterative performance. At smaller fill levels, the
sensitivity to the ILUT drop tolerance is more significant. For all
fill levels used, convergence was not achieved for the fuel assembly
problem with a drop tolerance larger than \sn{1}{-3}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge an
      eigenvalue iteration for the fuel assembly problem with ILUT
      preconditioning as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the iteration behavior for a
      different ILUT fill level. Fill levels of 1.5, 2.0, 2.5, and 3.0
      were used.}}
  \label{fig:ilut_iterations}
\end{figure}

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. Second, the matrix-matrix multiply operations required to
build the composite operator see significant performance losses due to
the amount of data that must be handled. In parallel, losing sparsity
severely inhibits performance where now parallel operations require
communications among orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.} \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
among states in the system and reduce the amount of memory used
along with the potential latency overhead. Given these two objectives,
the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/rda_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge a
      single eigenvalue iteration for the fuel assembly problem with
      each preconditioning as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4.25in]{chapters/spn_equations/rda_quality.pdf}
  \end{center}
  \caption{\textbf{Preconditioning quality metric for the fuel
      assembly problem given as a function of reduced domain
      approximation fill level.} \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_quality}
\end{figure}

@@

In this chapter, the Monte Carlo Synthetic Acceleration method has
been applied to the $SP_N$ discretization of the neutron transport
equation and explored in the context of a difficult nuclear fuel
assembly criticality problem. The following are the significant
observations and findings.
